---
title: "README"
output:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---

<!----------------------------------------------------------------------------->
<!----------------------------------------------------------------------------->
# neo2R

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/neo2R)](https://cran.r-project.org/package=neo2R)
[![](http://cranlogs.r-pkg.org/badges/neo2R)](https://cran.r-project.org/package=neo2R)

The aim of the neo2R is to provide simple and low level connectors
for querying [Neo4j graph databases](https://neo4j.com/).
The objects returned by the query functions are either lists or data.frames
with very few post-processing.
It allows fast processing of queries returning many records.
And it let the user handle post-processing according to the data model
and his needs.
It has been developed to support the BED package (https://github.com/patzaw/BED,
https://f1000research.com/articles/7-195/v3 ).
Other packages such as RNeo4j (https://github.com/nicolewhite/RNeo4j) or
neo4R (https://github.com/neo4j-rstats/neo4r) provide connectors to neo4j
databases with additional features.

<!----------------------------------------------------------------------------->
<!----------------------------------------------------------------------------->
# Installation

## From CRAN

```{r, eval=FALSE}
install.packages("neo2R")
```


<!------------------------->
## Dependencies

The following R packages available on CRAN are required:

    - base64enc
    - jsonlite
    - RCurl
    
They can be easily installed with the `install.packages()` function.

<!------------------------->
## Installation from github

```r
devtools::install_github("patzaw/neo2R")
```

<!----------------------------------------------------------------------------->
<!----------------------------------------------------------------------------->
# Use

<!------------------------->
## Running Neo4j

You can download and install Neo4j according to
the [documentation](https://neo4j.com/docs/getting-started/current/get-started-with-neo4j/#_installing_neo4j).
You can also run it in
a [docker container](https://neo4j.com/docs/operations-manual/current/docker/).
It takes a few seconds to start.

```{sh, eval=FALSE}
#!/bin/sh

#################################
## CONFIG according to your needs
#################################

export CONTAINER=neo4j_cont

## Chose Neo4j version (Only versions 3 and 4 are supported)
export NJ_VERSION=4.0.0
# export NJ_VERSION=3.5.14

## Ports
export NJ_HTTP_PORT=7474
export NJ_BOLT_PORT=7687

## Change the location of the Neo4j directory
export NJ_HOME=~/neo4j_home

## Authorization
NJ_AUTH=neo4j/1234 # set to 'none' if you want to disable authorization

## APOC download
export NJ_APOC=https://github.com/neo4j-contrib/neo4j-apoc-procedures/releases/download/4.0.0.2/apoc-4.0.0.2-all.jar
# export NJ_APOC=https://github.com/neo4j-contrib/neo4j-apoc-procedures/releases/download/3.5.0.7/apoc-3.5.0.7-all.jar

#################################
## RUN
#################################

mkdir -p $NJ_HOME

## Import and data directory
export NJ_IMPORT=$NJ_HOME/neo4jImport
mkdir -p $NJ_IMPORT
export NJ_DATA=$NJ_HOME/neo4jData
if test -e $NJ_DATA; then
   echo "$NJ_DATA directory exists ==> abort - Remove it before proceeding" >&2
   exit
fi
mkdir -p $NJ_DATA
## Neo4j plugins: APOC
export NJ_PLUGINS=$NJ_HOME/neo4jPlugins
mkdir -p $NJ_PLUGINS
cd $NJ_PLUGINS
wget --no-check-certificate $NJ_APOC
cd -

docker run -d \
   --name $CONTAINER \
   --publish=$NJ_HTTP_PORT:7474 \
   --publish=$NJ_BOLT_PORT:7687 \
   --env=NEO4J_dbms_memory_heap_initial__size=4G \
   --env=NEO4J_dbms_memory_heap_max__size=4G \
   --env=NEO4J_dbms_memory_pagecache_size=2G \
   --env=NEO4J_dbms_query__cache__size=0 \
   --env=NEO4J_cypher_min__replan__interval=100000000ms \
   --env=NEO4J_cypher_statistics__divergence__threshold=1 \
   --env=NEO4J_dbms_security_procedures_unrestricted=apoc.\\\* \
   --env=NEO4J_dbms_directories_import=import \
   --env NEO4J_AUTH=$NJ_AUTH \
   --volume $NJ_IMPORT:/var/lib/neo4j/import \
   --volume $NJ_DATA/data:/data \
	--volume $NJ_PLUGINS:/plugins \
	neo4j:$NJ_VERSION
	
```

<!------------------------->
## Connect to Neo4j

After installing [neo4j](https://neo4j.com/),
`startGraph` is used to initialize the connection from R.
If authentication has been disabled in neo4j by setting NEO4J.AUTH=none,
neither username nor password are required.
If you're connecting to a local instance of Neo4j and import directory has
been defined in the configuration, you can specify it in order to allow
import from data.frames.

```{r}
library(neo2R)
graph <- startGraph(
  "localhost:7474",
  username="neo4j", password="1234",
  importPath="~/neo4j_home/neo4jImport"
)
```

<!------------------------->
## Import from data.frame

If you're connected to a local instance of Neo4j and import directory has
been defined (see above), you can import data from data.frames.
Use the 'row' prefix to refer to the data.frame column.

```{r}
#########################################
## Nodes
## Create an index to speed-up MERGE
try(cypher(graph, 'CREATE INDEX ON :TestNode(name)'))
## Define node properties in a data.frame
set.seed(1)
nn <- 100000
nodes <- data.frame(
   "name"=paste(
      sample(LETTERS, nn, replace=TRUE),
      sample.int(nn, nn, replace=FALSE)
   ),
   "value"=rnorm(nn, 10, 3),
   stringsAsFactors=FALSE
)
import_from_df(
  graph=graph,
  cql='MERGE (n:TestNode {name:row.name, value:toFloat(row.value)})',
  toImport=nodes
)

#########################################
## Edges
## Define node properties in a data.frame
ne <- 100000
edges <- data.frame(
  "from"=sample(nodes$name, ne, replace=TRUE),
  "to"=sample(nodes$name, ne, replace=TRUE),
  "property"=round(runif(ne)*10),
  stringsAsFactors=FALSE
)
import_from_df(
   graph=graph,
   cql=prepCql(
      'MATCH (f:TestNode {name:row.from})',
      'MATCH (t:TestNode {name:row.to})',
      'MERGE (f)-[r:TestEdge {property:toInteger(row.property)}]->(t)'
   ),
   toImport=edges
)
```

<!------------------------->
## Query the Neo4j database

You can query the Neo4j graph database using the `cypher()` function.
Depending on the query, the function can return data in a a data.frame
(by setting `result="row"`) or in a list with nodes, relationships and paths
returned by the query (by setting `result="graph"`)

```{r}
## Get TestNode with value smaller than 4.
## According to the normal distribution we expect 2.5% of the total
## number of nodes ==> ~2500 nodes
df <- cypher(
   graph,
   prepCql(
      'MATCH (n:TestNode) WHERE n.value <= 4',
      'RETURN n.name as name, n.value as value'
   )
)
print(dim(df))
print(head(df))

## Get all paths of length 5 starting from a subset of nodes 
net <- cypher(
   graph,
   prepCql(
      'MATCH p=(f:TestNode)-[:TestEdge*5..5]->(t:TestNode) WHERE f.value < 3',
      'RETURN p'
   ),
   result="graph"
)
print(lapply(net, head, 3))
print(table(unlist(lapply(net$paths, length))))
```

