---
title: "Using neo2R with Local Neo4j Instances"
oauthor: "[Patrice Godard](mailto:patrice.godard@gmail.com?Subject=neo2R)"
date: "`r format(Sys.time(), '%B %d %Y')`"
always_allow_html: true
output:
  html_document:
    keep_md: no
    number_sections: yes
    self_contained: yes
    theme: cerulean
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
    number_sections: yes
package: neo2R (version `r packageVersion('neo2R')`)
vignette: >
   %\VignetteIndexEntry{Using neo2R with Local Neo4j Instances}
   %\VignetteEncoding{UTF-8}
   %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
--- 

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  echo = TRUE
)
```

# Introduction

This vignette demonstrates how to connect to a local Neo4j instance and import data from data frames.

```{r setup, include = FALSE}
# Check if we're running as pgodard and if the graph is accessible and empty
library(neo2R)

# Only run executable code if we're pgodard
is_pgodard <- identical(Sys.getenv("USER"), "pgodard")

# Try to connect to the local graph
if (is_pgodard) {
  graph <- tryCatch({
    startGraph(
      "https://localhost:7473",
      username = "neo4j",
      password = "donttrustusers",
      importPath = "~/neo4j_home/neo4jImport",
      .opts = list(ssl_verifypeer = 0)
    )
  }, error = function(e) NULL)
  
  # Check if graph is accessible
  should_run <- !is.null(graph)
} else {
  should_run <- FALSE
}
```

# Connecting to Local Neo4j

After installing [Neo4j](https://neo4j.com/), use `startGraph()` to initialize the connection from R. If authentication has been disabled in Neo4j by setting `NEO4J_AUTH=none`, neither username nor password are required.

If you're connecting to a local instance of Neo4j and an import directory has been defined in the configuration, you can specify it in order to allow import from data frames.

```{r, eval = should_run}
library(neo2R)

graph <- startGraph(
  "https://localhost:7473",
  username = "neo4j",
  password = "donttrustusers",
  importPath = "~/neo4j_home/neo4jImport",
  .opts = list(ssl_verifypeer = 0)
)
```

# Importing Data from Data Frames

If you're connected to a local instance of Neo4j and an import directory has been defined (see above), you can import data from data frames. Use the `row` prefix to refer to the data frame column.

```{r, eval = should_run}
# Create an index to speed-up MERGE
if (as.integer(graph$version[[1]]) >= 5) {
   try(cypher(graph, 'CREATE INDEX FOR (n:TestNode) ON (n.name)'), silent = TRUE)
} else {
   try(cypher(graph, 'CREATE INDEX ON :TestNode(name)'), silent = TRUE)
}

# Define node properties in a data frame
set.seed(1)
nn <- 100000
nodes <- data.frame(
   "name" = paste(
      sample(LETTERS, nn, replace = TRUE),
      sample.int(nn, nn, replace = FALSE)
   ),
   "value" = rnorm(nn, 10, 3),
   stringsAsFactors = FALSE
)

# Import nodes
import_from_df(
  graph = graph,
  cql = 'MERGE (n:TestNode {name: row.name, value: toFloat(row.value)})',
  toImport = nodes
)

# Define edge properties in a data frame
ne <- 100000
edges <- data.frame(
  "from" = sample(nodes$name, ne, replace = TRUE),
  "to" = sample(nodes$name, ne, replace = TRUE),
  "property" = round(runif(ne) * 10),
  stringsAsFactors = FALSE
)

# Import edges
import_from_df(
   graph = graph,
   cql = prepCql(
      'MATCH (f:TestNode {name: row.from})',
      'MATCH (t:TestNode {name: row.to})',
      'MERGE (f)-[r:TestEdge {property: toInteger(row.property)}]->(t)'
   ),
   toImport = edges
)
```

# Querying the Neo4j Database

You can query the Neo4j graph database using the `cypher()` function. Depending on the query, the function can return data in a data frame (by setting `result = "row"`) or in a list with nodes, relationships and paths returned by the query by setting `result = "graph"`.

```{r, eval = should_run}
# Get TestNode with value smaller than 4
# According to the normal distribution we expect 2.5% of the total
# number of nodes ==> ~2500 nodes
df <- cypher(
   graph,
   prepCql(
      'MATCH (n:TestNode) WHERE n.value <= 4',
      'RETURN n.name as name, n.value as value'
   )
)
print(dim(df))
print(head(df))

# Multiple queries can be sent at once
dfl <- multicypher(
   graph,
   sprintf(
      paste(
         'MATCH (n:TestNode) WHERE n.value <= %s',
         'RETURN n.name as name, n.value as value'
      ),
      2:4
   )
)
print(lapply(dfl, dim))

# Get all paths of length 5 starting from a subset of nodes 
net <- cypher(
   graph,
   prepCql(
      'MATCH p=(f:TestNode)-[:TestEdge*5..5]->(t:TestNode) WHERE f.value < 3',
      'RETURN p'
   ),
   result = "graph"
)
print(lapply(net, head, 3))
print(table(unlist(lapply(net$paths, length))))
```

## Further Reading

- [Neo4j Cypher reference](https://neo4j.com/docs/cypher-manual/current/) — query language docs
