---
title: "Running Neo4j with Docker Containers"
author: "[Patrice Godard](mailto:patrice.godard@gmail.com?Subject=neo2R)"
date: "`r format(Sys.time(), '%B %d %Y')`"
always_allow_html: true
output:
  html_document:
    keep_md: no
    number_sections: yes
    self_contained: yes
    theme: cerulean
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
    number_sections: yes
package: neo2R (version `r packageVersion('neo2R')`)
vignette: >
   %\VignetteIndexEntry{Running Neo4j with Docker Containers}
   %\VignetteEncoding{UTF-8}
   %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
--- 

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  echo = TRUE
)
```

# Introduction

This vignette shows how to instantiate Docker containers running different versions of Neo4j (3.x, 4.x, 5.x, and 2025+). The main differences between the configurations are related to the APOC library and SSL setup.

# Prerequisites

1. Install Docker on your system
2. Install openssl for SSL certificate generation

# Neo4j 3.x

For Neo4j 3.x, you need to manually download and configure the APOC library.

```{sh, eval=FALSE}
#!/bin/sh

#################################
## CONFIG according to your needs
#################################

export CONTAINER=neo4j_cont
export NJ_VERSION=3.5.30

## Ports
export NJ_HTTP_PORT=7474
export NJ_HTTPS_PORT=7473
export NJ_BOLT_PORT=7687

## Change the location of the Neo4j directory
export NJ_HOME=~/neo4j_home

## Authorization
## set to 'none' if you want to disable authorization
NJ_AUTH=neo4j/donttrustusers

## APOC download
export NJ_APOC=https://github.com/neo4j-contrib/neo4j-apoc-procedures/releases/download/3.5.0.15/apoc-3.5.0.15-all.jar

#################################
## RUN
#################################

mkdir -p $NJ_HOME

## Import and data directory
export NJ_IMPORT=$NJ_HOME/neo4jImport
mkdir -p $NJ_IMPORT
export NJ_DATA=$NJ_HOME/neo4jData
if test -e $NJ_DATA; then
   echo "$NJ_DATA directory exists ==> abort - Remove it before proceeding" >&2
   exit
fi
mkdir -p $NJ_DATA

## SSL
export NJ_SSL=${NJ_HOME}/ssl
mkdir -p ${NJ_SSL}/client_policy
mkdir -p ${NJ_SSL}/client_policy/revoked
mkdir -p ${NJ_SSL}/client_policy/trusted
openssl req -subj "/CN=localhost" -new -newkey rsa:2048 \
   -days 365 -nodes -x509 \
   -keyout ${NJ_SSL}/client_policy/private.key \
   -out ${NJ_SSL}/client_policy/public.crt
chmod -R a+rwx ${NJ_SSL}

## Neo4j plugins: APOC
export NJ_PLUGINS=$NJ_HOME/neo4jPlugins
mkdir -p $NJ_PLUGINS
cd $NJ_PLUGINS
wget --no-check-certificate $NJ_APOC
cd -

docker run -d \
   --name $CONTAINER \
   --publish=$NJ_HTTP_PORT:7474 \
   --publish=$NJ_HTTPS_PORT:7473 \
   --publish=$NJ_BOLT_PORT:7687 \
   --env=NEO4J_dbms_memory_heap_initial__size=2G \
   --env=NEO4J_dbms_memory_heap_max__size=2G \
   --env=NEO4J_dbms_memory_pagecache_size=1G \
   --env=NEO4J_dbms_query__cache__size=0 \
   --env=NEO4J_cypher_min__replan__interval=100000000ms \
   --env=NEO4J_cypher_statistics__divergence__threshold=1 \
   --env=NEO4J_dbms_security_procedures_unrestricted=apoc.\\* \
   --env=NEO4J_dbms_directories_import=import \
   --env NEO4J_AUTH=$NJ_AUTH \
   --volume $NJ_IMPORT:/var/lib/neo4j/import \
   --volume $NJ_DATA/data:/data \
   --volume $NJ_PLUGINS:/plugins \
   --volume=$NJ_SSL:/ssl \
   --env=NEO4J_https_ssl__policy=client_policy \
   --env=NEO4J_dbms_ssl_policy_client__policy_base__directory=/ssl/client_policy \
   --env=NEO4J_dbms_ssl_policy_client__policy_client__auth=NONE \
   --env=NEO4J_dbms_ssl_policy_client__policy_trust__all=true \
   neo4j:$NJ_VERSION

sleep 15
sudo chmod a+rwx $NJ_IMPORT
```

# Neo4j 4.x

Neo4j 4.x has similar configuration to 3.x but with updated environment variable names.

```{sh, eval=FALSE}
#!/bin/sh

#################################
## CONFIG according to your needs
#################################

export CONTAINER=neo4j_cont
export NJ_VERSION=4.4.26

## Ports
export NJ_HTTP_PORT=7474
export NJ_HTTPS_PORT=7473
export NJ_BOLT_PORT=7687

## Change the location of the Neo4j directory
export NJ_HOME=~/neo4j_home

## Authorization
export NJ_AUTH=neo4j/donttrustusers

## APOC download
export NJ_APOC=https://github.com/neo4j-contrib/neo4j-apoc-procedures/releases/download/4.4.0.3/apoc-4.4.0.3-all.jar

#################################
## RUN
#################################

mkdir -p $NJ_HOME

## Import and data directory
export NJ_IMPORT=$NJ_HOME/neo4jImport
mkdir -p $NJ_IMPORT
export NJ_DATA=$NJ_HOME/neo4jData
if test -e $NJ_DATA; then
   echo "$NJ_DATA directory exists ==> abort - Remove it before proceeding" >&2
   exit
fi
mkdir -p $NJ_DATA

## SSL
export NJ_SSL=${NJ_HOME}/ssl
mkdir -p ${NJ_SSL}/https
mkdir -p ${NJ_SSL}/https/revoked
mkdir -p ${NJ_SSL}/https/trusted
openssl req -subj "/CN=localhost" -new -newkey rsa:2048 \
   -days 365 -nodes -x509 \
   -keyout ${NJ_SSL}/https/private.key \
   -out ${NJ_SSL}/https/public.crt
chmod -R a+rwx ${NJ_SSL}

## Neo4j plugins: APOC
export NJ_PLUGINS=$NJ_HOME/neo4jPlugins
mkdir -p $NJ_PLUGINS
cd $NJ_PLUGINS
wget --no-check-certificate $NJ_APOC
cd -

docker run -d \
   --name $CONTAINER \
   --publish=$NJ_HTTP_PORT:7474 \
   --publish=$NJ_HTTPS_PORT:7473 \
   --publish=$NJ_BOLT_PORT:7687 \
   --env=NEO4J_dbms_memory_heap_initial__size=2G \
   --env=NEO4J_dbms_memory_heap_max__size=2G \
   --env=NEO4J_dbms_memory_pagecache_size=1G \
   --env=NEO4J_dbms_query__cache__size=0 \
   --env=NEO4J_cypher_min__replan__interval=100000000ms \
   --env=NEO4J_cypher_statistics__divergence__threshold=1 \
   --env=NEO4J_dbms_security_procedures_unrestricted=apoc.\\* \
   --env=NEO4J_dbms_directories_import=import \
   --env NEO4J_AUTH=$NJ_AUTH \
   --volume $NJ_IMPORT:/var/lib/neo4j/import \
   --volume $NJ_DATA/data:/data \
   --volume $NJ_PLUGINS:/plugins \
   --volume=$NJ_SSL:/ssl \
   --env=NEO4J_dbms_connector_https_enabled=true \
   --env=NEO4J_dbms_ssl_policy_https_enabled=true \
   --env=NEO4J_dbms_ssl_policy_https_base__directory=/ssl/https \
   --env=NEO4J_dbms_ssl_policy_https_private__key=private.key \
   --env=NEO4J_dbms_ssl_policy_https_public__certificate=public.crt \
   --env=NEO4J_dbms_ssl_policy_https_client__auth=NONE \
   --env=NEO4J_dbms_ssl_policy_https_trust__all=true \
   neo4j:$NJ_VERSION

sleep 15
sudo chmod a+rwx $NJ_IMPORT
```

# Neo4j 5.x (>= 5.12)

In Neo4j 5.x, it is not possible to instantiate the Docker image with a specific username and password. The password must be changed after Neo4j starts. However, you can instantiate the image without any credentials using `--env NEO4J_AUTH=none`.

```{sh, eval=FALSE}
#!/bin/sh

#################################
## CONFIG according to your needs
#################################

export CONTAINER=neo4j_cont
export NJ_VERSION=5.26.26

## Ports
export NJ_HTTP_PORT=7474
export NJ_HTTPS_PORT=7473
export NJ_BOLT_PORT=7687

## Change the location of the Neo4j directory
export NJ_HOME=~/neo4j_home

## APOC download
export NJ_APOC=https://github.com/neo4j/apoc/releases/download/5.26.26/apoc-5.26.26-core.jar

#################################
## RUN
#################################

mkdir -p $NJ_HOME

## Import and data directory
export NJ_IMPORT=$NJ_HOME/neo4jImport
mkdir -p $NJ_IMPORT
export NJ_DATA=$NJ_HOME/neo4jData
if test -e $NJ_DATA; then
   echo "$NJ_DATA directory exists ==> abort - Remove it before proceeding" >&2
   exit
fi
mkdir -p $NJ_DATA
export NJ_LOGS=$NJ_HOME/neo4jLogs
mkdir -p $NJ_LOGS

## SSL
export NJ_SSL=${NJ_HOME}/ssl
mkdir -p ${NJ_SSL}/https
mkdir -p ${NJ_SSL}/https/revoked
mkdir -p ${NJ_SSL}/https/trusted
openssl req -subj "/CN=localhost" -new -newkey rsa:2048 \
   -days 365 -nodes -x509 \
   -keyout ${NJ_SSL}/https/private.key \
   -out ${NJ_SSL}/https/public.crt
cp ${NJ_SSL}/https/public.crt ${NJ_SSL}/https/trusted
chmod -R a+rwx ${NJ_SSL}

## Neo4j plugins: APOC
export NJ_PLUGINS=$NJ_HOME/neo4jPlugins
mkdir -p $NJ_PLUGINS
cd $NJ_PLUGINS
wget --no-check-certificate $NJ_APOC
cd -

## Add ` --env NEO4J_AUTH=none ` if you don't want to setup credentials
docker run -d \
   --name $CONTAINER \
   --publish=$NJ_HTTP_PORT:7474 \
   --publish=$NJ_HTTPS_PORT:7473 \
   --publish=$NJ_BOLT_PORT:7687 \
   --env=NEO4J_server_memory_heap_initial__size=2G \
   --env=NEO4J_server_memory_heap_max__size=2G \
   --env=NEO4J_server_memory_pagecache_size=1G \
   --env=NEO4J_server_db_query__cache__size=0 \
   --env=NEO4J_dbms_cypher_min__replan__interval=100000000ms \
   --env=NEO4J_dbms_cypher_statistics__divergence__threshold=1 \
   --env=NEO4J_dbms_security_procedures_unrestricted=apoc.\\* \
   --env=NEO4J_server_directories_import=import\
   --volume=$NJ_IMPORT:/var/lib/neo4j/import \
   --volume=$NJ_DATA/data:/data \
   --volume=$NJ_LOGS:/var/lib/neo4j/logs \
   --volume=$NJ_PLUGINS:/plugins \
   --volume=$NJ_SSL:/ssl \
   --env=NEO4J_server_https_enabled=true \
   --env=NEO4J_dbms_ssl_policy_https_enabled=true \
   --env=NEO4J_dbms_ssl_policy_https_base__directory=/ssl/https \
   --env=NEO4J_dbms_ssl_policy_https_private__key=private.key \
   --env=NEO4J_dbms_ssl_policy_https_public__certificate=public.crt \
   --env=NEO4J_dbms_ssl_policy_https_client__auth=NONE \
   --env=NEO4J_dbms_ssl_policy_https_trust__all=true \
   neo4j:$NJ_VERSION

sleep 15
sudo chmod a+rwx $NJ_IMPORT
```

# Neo4j >= 2025

In recent versions of Neo4j (2025 and later), the APOC jar is bundled in the image. The `NEO4J_PLUGINS` environment variable automatically copies it in at container start, and persisting `/plugins` avoids re-copying on every restart.

```{sh, eval=FALSE}
#!/bin/sh

#################################
## CONFIG according to your needs
#################################

export CONTAINER=neo4j_cont
export NJ_VERSION=2026.06.0

## Ports
export NJ_HTTP_PORT=7474
export NJ_HTTPS_PORT=7473
export NJ_BOLT_PORT=7687

## Change the location of the Neo4j directory
export NJ_HOME=~/neo4j_home

#################################
## RUN
#################################

mkdir -p $NJ_HOME

## Import and data directory
export NJ_IMPORT=$NJ_HOME/neo4jImport
mkdir -p $NJ_IMPORT
export NJ_DATA=$NJ_HOME/neo4jData
if test -e $NJ_DATA; then
   echo "$NJ_DATA directory exists ==> abort - Remove it before proceeding" >&2
   exit
fi
mkdir -p $NJ_DATA
export NJ_LOGS=$NJ_HOME/neo4jLogs
mkdir -p $NJ_LOGS

## SSL
export NJ_SSL=${NJ_HOME}/ssl
mkdir -p ${NJ_SSL}/https
mkdir -p ${NJ_SSL}/https/revoked
mkdir -p ${NJ_SSL}/https/trusted
openssl req -subj "/CN=localhost" -new -newkey rsa:2048 \
   -days 365 -nodes -x509 \
   -keyout ${NJ_SSL}/https/private.key \
   -out ${NJ_SSL}/https/public.crt
cp ${NJ_SSL}/https/public.crt ${NJ_SSL}/https/trusted
chmod -R a+rwx ${NJ_SSL}

## Plugins directory (APOC jar is now bundled in the image;
## NEO4J_PLUGINS copies it in automatically at container start,
## and persisting /plugins avoids re-copying on every restart)
export NJ_PLUGINS=$NJ_HOME/neo4jPlugins
mkdir -p $NJ_PLUGINS

## Add ` --env NEO4J_AUTH=none ` if you don't want to setup credentials
docker run -d \
   --name $CONTAINER \
   --publish=$NJ_HTTP_PORT:7474 \
   --publish=$NJ_HTTPS_PORT:7473 \
   --publish=$NJ_BOLT_PORT:7687 \
   --env=NEO4J_server_memory_heap_initial__size=2G \
   --env=NEO4J_server_memory_heap_max__size=2G \
   --env=NEO4J_server_memory_pagecache_size=1G \
   --env=NEO4J_server_db_query__cache__size=0 \
   --env=NEO4J_dbms_cypher_min__replan__interval=100000000ms \
   --env=NEO4J_dbms_cypher_statistics__divergence__threshold=1 \
   --env=NEO4J_PLUGINS='["apoc"]' \
   --env=NEO4J_dbms_security_procedures_unrestricted=apoc.\\* \
   --env=NEO4J_server_directories_import=import\
   --volume=$NJ_IMPORT:/var/lib/neo4j/import \
   --volume=$NJ_DATA/data:/data \
   --volume=$NJ_LOGS:/var/lib/neo4j/logs \
   --volume=$NJ_PLUGINS:/plugins \
   --volume=$NJ_SSL:/ssl \
   --env=NEO4J_server_https_enabled=true \
   --env=NEO4J_dbms_ssl_policy_https_enabled=true \
   --env=NEO4J_dbms_ssl_policy_https_base__directory=/ssl/https \
   --env=NEO4J_dbms_ssl_policy_https_private__key=private.key \
   --env=NEO4J_dbms_ssl_policy_https_public__certificate=public.crt \
   --env=NEO4J_dbms_ssl_policy_https_client__auth=NONE \
   --env=NEO4J_dbms_ssl_policy_https_trust__all=true \
   neo4j:$NJ_VERSION

sleep 15
sudo chmod a+rwx $NJ_IMPORT
```

# Updating Credentials for Neo4j >= 5

If you did not disable credentials (with `--env NEO4J_AUTH=none`), you'll need to first change the user password before being able to connect to the graph database. The following code shows how to do it with neo2R:

```{r, eval=FALSE}
library(neo2R)

# First connect with default credentials
system <- startGraph(
  "https://localhost:7473",
  database = "system",
  check = FALSE,
  username = "neo4j",
  password = "neo4j",
  importPath = "~/neo4j_home/neo4jImport",
  .opts = list(ssl_verifypeer = 0)
)

# Change the password
cypher(
  system,
  "ALTER CURRENT USER SET PASSWORD FROM 'neo4j' TO 'donttrustusers'"
)
```

# Key Differences Summary

| Version | APOC Configuration | SSL Configuration | Authentication |
|---------|-------------------|-------------------|----------------|
| 3.x | Manual download, mount as volume | `client_policy` directory | Set via `NEO4J_AUTH` |
| 4.x | Manual download, mount as volume | `https` directory | Set via `NEO4J_AUTH` |
| 5.x | Manual download, mount as volume | `https` directory | Change after start or use `NEO4J_AUTH=none` |
| 2025+ | Bundled, use `NEO4J_PLUGINS` | `https` directory | Change after start or use `NEO4J_AUTH=none` |

## Further Reading

- [Neo4j Docker documentation](https://neo4j.com/docs/operations-manual/current/docker/) — Running Neo4j in a Docker container.
